\section{Parallel \texorpdfstring{\MS{}}{MergeSort}}
\label{sec:mram:par}

A simplistic way to parallelise \MS{} is the following:
Let the number of tasklets be a power of two and have the tasklet identifiers start from zero.
The whole input array is divided into as many shares of equal length as there are tasklets, and each tasklet sorts a share sequentially using the \ac{MRAM} \MS{} of \cref{sec:mram:merge} to form starting runs.
Once finished, Tasklet~\(t\) with~\(t \bmod 2 = 1\) informs Tasklet~\(t - 1\) that it is finished with sorting its share and suspends itself.
Tasklet \(t - 1\) merges its own share and that of Tasklet~\(t\) into a bigger run.
Once finished, Tasklet~\(t\) with~\(t \bmod 4 = 2\) informs Tasklet~\(t - 2\) that it is finished with sorting the run and suspends itself.
Then, Tasklet~\(t - 2\) merges its run with that of Tasklet~\(t\).
This scheme continues until the last round where the two remaining runs are sorted by Tasklet 0.

The bottleneck is the sequential execution of each merge which eventually leads to a single active tasklet.
Even with infinite many processors, this simplistic parallel \MS{} can achieve a theoretical parallel speedup\footnote{
	The \emph{parallel speedup} \(S = \operatorname{t}_1(A) / \operatorname{t}_p(A)\) of a parallel algorithm \(A\) is the ratio of its wall-clock time \(\operatorname{t}_1(A)\) when run with one processor and of its wall-clock time \(\operatorname{t}_p(A)\) when run with \(p\) processors.
} of at most \(\bigtheta{\log n}\).
\Citeauthor{cormen2013algorithmen}~\cite{cormen2013algorithmen} propose an alternative approach whose maximum theoretical parallel speedup is \(\bigtheta{n / \log^2 n}\).
Advantageously, \cref{alg:mram:two-tier merge} for merging \ac{MRAM} data can be reused without fundamental changes when adapting this parallel \MS{} to \acp{DPU}.
Also, the number of synchronisation points is logarithmic in the number of tasklets only, and the time which each synchronisation takes is insignificant compared to the total runtime.

\input{mram_par_aspects}

\input{mram_par_performance}
