\subsection{\texorpdfstring{\IS{}}{InsertionSort}}
\label{subsec:tasklet:insertion}

This stable sorting algorithm works by moving the \(i\)th element leftwards as long as its left neighbour is greater, assuming that the elements at the indices \(0\) to \(i - 1\) are already sorted.
Its asymptotic runtime is bad, reaching \(\bigoh*{n^2}\) not only in the worst case but also in the average case, where each of the \(\binom{n}{2}\) pairs of input elements is in wrong order and will be swapped at some point in the execution with probability 50\%.
Nonetheless, \IS{} does have some saving graces:
\begin{enumerate*}
	\item
	If the input array is mostly or even fully sorted, the runtime drops down to \(\bigoh{n}\).

	\item
	It works in-place, needing only \(\bigoh{1}\) additional space.

	\item
	Its program code is short, lending itself to inlining.

	\item
	The overhead is small.
\end{enumerate*}
Especially the last two points make \IS{} a good fall-back algorithm for asymptotically better sorting algorithms to use on short subarrays.

\paragraph{Sentinel Values}
When moving an element to the left, two checks are needed:
Does the left neighbour exist and is it smaller than the element to move?
The first check can be omitted through the use of \emph{sentinel values}:
If the element at index \(-1\) is the smallest possible value of the chosen data type, it is at least as small as any value in the input array, and the leftwards motion stops there at the latest.
Since a DPU has no branch predictor, the slowdown from performing twice as many checks as needed is quite high and lies at about 30\% for short inputs (\cref{fig:insertion:against_others}).

Setting such an \emph{explicit} sentinel value can be omitted by using \emph{implicit} sentinel values.
At the start of each round, one can check if the element at index \(0\) is at least as small as the element at index~\(i\).
If yes, the former is a sufficient sentinel value, and \IS{} can proceed as normal.
If not, the latter must be the minimum of the first \(i + 1\) elements and, therefore, can be moved immediately to the front.

\input{tasklet_insertion_performance.tex}

\input{tasklet_insertion_compilation}
